Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-08-15 ident_manu.Rnw
added calex figure from new flask calibration data

@@

cap_val <- 'Results of model calibration to observed \\ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\\gamma$) and parameters of each subset are from \\cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.'
foot.val <- '\\footnotesize *Starting \\ac{rmse} $= 72.140$'
colnms <- c('$\\gamma$', '$n$ parameters', 'mean L1', 'Iterations', '$RMSE$*')

@@

<<areachg, eval = T, fig.cap = 'Estimated hypoxic area and monthly averages (km$^2 \\times$ 1000) using the default parameter settings (a, b) and the differences using a 50\\% increase in each parameter (c, d).  The selected parameters were the top sensitive in each parameter category for \\ac{do} in \\cref{fig:sensalltile,tab:dosens}. Error bars are 95\\% confidence intervals for the monthly averages in (b) and the difference in monthly averages (d) from the perturbed and default results.  Areal extent of estimated hypoxia on August 28\\textsuperscript{th} using the default parameters (grey) and estimated changes with parameter perturbations is shown in (e).', fig.height = 7, fig.width = 7, echo = F, cache = TRUE, out.width = "\\textwidth">>=
# ##
# # area time series
# areats <- read_excel('ignore/runcompare.xlsx', sheet = 'area')
# 
# ##
# # get all arrays of hypox area for each parameter
# shts <- excel_sheets('ignore/runcompare.xlsx') %>% 
#   grep('array', ., value = T)
# latlon <- read_excel('ignore/runcompare.xlsx', sheet = 'Latlon') %>% 
#   list(
#     lat = c(.[,1])[[1]], 
#     lon = as.numeric(names(.)[-1])
#     ) %>% 
#   .[c('lat', 'lon')]
# 
# ls <- vector('list', length = length(shts))
# names(ls) <- shts
# for(sht in shts){
#  
#   cat(sht, '\t')
#   
#   # read the sheet, interpolate for higher res
#   tmp <- read_excel('ignore/runcompare.xlsx', sheet = sht, na = '-9999', col_names = F) 
#   tmp <- tmp[1:176,1:320] %>% 
#     as.matrix
#   obs <- list(x = latlon$lat, y = latlon$lon, z = tmp)
#   newgrd <- expand.grid(
#       x = approx(latlon$lat, n = 4 * ncol(tmp))$y,
#       y = approx(latlon$lon, n = 4 * nrow(tmp))$y
#       )
#   int_val <- fields::interp.surface(
#     obj = obs,
#     loc = newgrd
#     )
#   tmp <- data.frame(newgrd, tmp = int_val) %>%
#     spread(x, tmp) %>%
#     dplyr::select(-y) %>%
#     as.matrix %>% 
#     .[, ncol(.):1] %>% 
#     t
#   
#   # subset hypoxic areas, make polygon
#   # get only the large areas
#   tmp[tmp >= 64] <- NA
#   tmp[!is.na(tmp)] <- 1
#   tmp <- raster(tmp, 
#     xmn = min(latlon$lon), 
#     xmx = max(latlon$lon),   
#     ymn = min(latlon$lat), 
#     ymx = max(latlon$lat)
#     )
#   tmp <- rasterToPolygons(tmp, dissolve = T) %>% 
#     disaggregate
#   areas <- transform(data.frame(tmp), area = gArea(tmp, byid = T))
#   filt <- with(areas, which(area > 0.12)) #quantile(area, 0.99)))
#   tmp <- tmp[filt, ] %>% 
#     fortify
#   
#   ls[[sht]] <- data.frame(parm = sht, tmp)
#    
# }
# 
# # combine output from each sheet
# arrs <- do.call('rbind', ls) %>% 
#   mutate(
#     parm = gsub('_array$', '', parm)
#     ) %>% 
#   # filter(lat > 28) %>% 
#   tibble::remove_rownames(.) %>% 
#   unite('grp', parm, group, remove = F)
# 
# # save
# save(areats, arrs, file = 'data/area.RData', compress = 'xz')

>>>>

load(file = 'data/cals.RData')

# starting rmse
strt <- cals[[1]]$mat[1,] %>% 
  .[length(.)] %>% 
  scinot(.)

@@

cap_val <- 'Results of model calibration to observed \\ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\\gamma$) and parameters of each subset are from \\cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.'
foot.val <- paste0('\\footnotesize *Starting \\ac{rmse} ', strt)
colnms <- c('$\\gamma$', '$n$ parameters', 'mean L1', 'Iterations', '$RMSE$*')

@@

% calibration example
<<calex, eval = T, fig.cap = 'Example of calibration steps for the phytoplankton parameter subset.  The FishTank model was calibrated using data from an approximate 12-hour flask experiment of oxygen production.  Model calibration adjusted parameters (left plots) to minimize the \\ac{rmse} between experimental and modelled dissolved oxygen (top left).  The phytoplankton parameters were those identifiable for dissolved oxygen (\\cref{tab:heurist1}), arranged from most (top plot) to least sensitive (bottom plot) (\\cref{tab:dosens}).  The relative parameter changes in the right plots show the initial parameter values (black) and the the selected parameter values (blue) after calibration.', fig.height = 7, fig.width = 6, echo = F, cache = T, out.width = '\\textwidth'>>=
data(cals)
data(tocal_all)
plo_cal(cals, tocal_all, whichsub = 'Phytoplankton', widths = c(1, 0.4), heights = c(0.125, 1))
@

% CGEM area change plots
<<areachg, eval = T, fig.cap = 'Estimated hypoxic area and monthly averages (km$^2 \\times$ 1000) using the default parameter settings (a, b) and the differences using a 50\\% increase in each parameter (c, d).  The selected parameters were the top sensitive in each parameter category for \\ac{do} in \\cref{fig:sensalltile,tab:dosens}. Error bars are 95\\% confidence intervals for the monthly averages in (b) and the difference in monthly averages (d) from the perturbed and default results.  Areal extent of estimated hypoxia on August 28\\textsuperscript{th} using the default parameters (grey) and estimated changes with parameter perturbations is shown in (e).', fig.height = 7, fig.width = 7, echo = F, cache = TRUE, out.width = "\\textwidth">>=
# ##
# # area time series
# areats <- read_excel('ignore/runcompare.xlsx', sheet = 'area')
# 
# ##
# # get all arrays of hypox area for each parameter
# shts <- excel_sheets('ignore/runcompare.xlsx') %>% 
#   grep('array', ., value = T)
# latlon <- read_excel('ignore/runcompare.xlsx', sheet = 'Latlon') %>% 
#   list(
#     lat = c(.[,1])[[1]], 
#     lon = as.numeric(names(.)[-1])
#     ) %>% 
#   .[c('lat', 'lon')]
# 
# ls <- vector('list', length = length(shts))
# names(ls) <- shts
# for(sht in shts){
#  
#   cat(sht, '\t')
#   
#   # read the sheet, interpolate for higher res
#   tmp <- read_excel('ignore/runcompare.xlsx', sheet = sht, na = '-9999', col_names = F) 
#   tmp <- tmp[1:176,1:320] %>% 
#     as.matrix
#   obs <- list(x = latlon$lat, y = latlon$lon, z = tmp)
#   newgrd <- expand.grid(
#       x = approx(latlon$lat, n = 4 * ncol(tmp))$y,
#       y = approx(latlon$lon, n = 4 * nrow(tmp))$y
#       )
#   int_val <- fields::interp.surface(
#     obj = obs,
#     loc = newgrd
#     )
#   tmp <- data.frame(newgrd, tmp = int_val) %>%
#     spread(x, tmp) %>%
#     dplyr::select(-y) %>%
#     as.matrix %>% 
#     .[, ncol(.):1] %>% 
#     t
#   
#   # subset hypoxic areas, make polygon
#   # get only the large areas
#   tmp[tmp >= 64] <- NA
#   tmp[!is.na(tmp)] <- 1
#   tmp <- raster(tmp, 
#     xmn = min(latlon$lon), 
#     xmx = max(latlon$lon),   
#     ymn = min(latlon$lat), 
#     ymx = max(latlon$lat)
#     )
#   tmp <- rasterToPolygons(tmp, dissolve = T) %>% 
#     disaggregate
#   areas <- transform(data.frame(tmp), area = gArea(tmp, byid = T))
#   filt <- with(areas, which(area > 0.12)) #quantile(area, 0.99)))
#   tmp <- tmp[filt, ] %>% 
#     fortify
#   
#   ls[[sht]] <- data.frame(parm = sht, tmp)
#    
# }
# 
# # combine output from each sheet
# arrs <- do.call('rbind', ls) %>% 
#   mutate(
#     parm = gsub('_array$', '', parm)
#     ) %>% 
#   # filter(lat > 28) %>% 
#   tibble::remove_rownames(.) %>% 
#   unite('grp', parm, group, remove = F)
# 
# # save
# save(areats, arrs, file = 'data/area.RData', compress = 'xz')

### 2017-08-15 ident_manu.tex
added calex figure from new flask calibration data

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Fri Aug 4 17:59:08 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/84e94b5eb0aca29df79344221e5651a5af428017}{84e94b5eb0aca29df79344221e5651a5af428017}\newline Acronyms: \acfi{chla}; \acfi{cgem}; \acfi{do}; \acfi{dom}; \acfi{gom}; \acfi{lcs}; \acfi{marb}; \acfi{pom}; \acfi{rmse}; \acfi{zerod}}

@@

Calibration of parameter values for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3})  produced model output that varied in goodness of fit and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only six iterations were required to optimize the single parameter selected for the temperature category, whereas 136 iterations were required for the eight parameters selected independent of category.  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, selection independent of parameter category and equally within each category produced the largest reduction in \ac{rmse} values from the default output (starting \ac{rmse} $33.47$ mmol O$_2$ m$^{-3}$, reduction to $26.56$, $28.36$, respectively), whereas the optimal values for parameters in the Optics category had the smallest reduction ($33.3$).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 ($2.81\times 10^{-4}$, three parameters), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 ($0.03$, eight parameters, \cref{tab:calib}).

@@

% calibration table, do
%latex.default(totab[, -1], file = "", rowlabel = "Selection heuristic",     caption = cap_val, caption.loc = "top", size = "scriptsize",     label = "tab:calib", rowname = totab$heurist, rgroup = c("1",         "2", "3"), insert.bottom = foot.val, n.rgroup = c(5,         1, 1), colheads = colnms)%
\begin{table}[!tbp]
{\scriptsize
\caption{Results of model calibration to observed \ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\gamma$) and parameters of each subset are from \cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.\label{tab:calib}} 
\begin{center}
\begin{tabular}{llllll}
\hline\hline
\multicolumn{1}{l}{Selection heuristic}&\multicolumn{1}{c}{$\gamma$}&\multicolumn{1}{c}{$n$ parameters}&\multicolumn{1}{c}{mean L1}&\multicolumn{1}{c}{Iterations}&\multicolumn{1}{c}{$RMSE$*}\tabularnewline
\hline
{\bfseries 1}&&&&&\tabularnewline
~~Optics&$3.87$&$3$&$2.81\times 10^{-4}$&$49$&$33.3$\tabularnewline
~~Organic Matter&$8.61$&$5$&$3.18\times 10^{-3}$&$44$&$31.48$\tabularnewline
~~Phytoplankton&$3.97$&$7$&$0.02$&$300$&$25.84$\tabularnewline
~~Temperature&$1$&$1$&$0.02$&$6$&$31.23$\tabularnewline
~~Zooplankton&$5.2$&$4$&$0.03$&$72$&$32.99$\tabularnewline
\hline
{\bfseries 2}&&&&&\tabularnewline
~~Independent of category&$6.81$&$8$&$0.03$&$136$&$26.56$\tabularnewline
\hline
{\bfseries 3}&&&&&\tabularnewline
~~Equally within category&$5.22$&$7$&$0.03$&$45$&$28.36$\tabularnewline
\hline
\end{tabular}\end{center}}
\footnotesize *Starting \ac{rmse} $= 72.140$\end{table}

@@

\begin{figure}[!ht]

>>>>

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Sun Aug 6 08:21:57 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/c03d0df749cea19daced8a68105265cf1ed2a63e}{c03d0df749cea19daced8a68105265cf1ed2a63e}\newline Acronyms: \acfi{chla}; \acfi{cgem}; \acfi{do}; \acfi{dom}; \acfi{gom}; \acfi{lcs}; \acfi{marb}; \acfi{pom}; \acfi{rmse}; \acfi{zerod}}

@@

Calibration of parameter values for each of the seven selection heuristics (\cref{tab:heurist1,tab:heurist2,tab:heurist3})  produced model output that varied in goodness of fit and computational effort required to identify optimal values (\cref{tab:calib}).  As expected, the number of required iterations of the optimization algorithm to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only three iterations were required to optimize the single parameter selected for the temperature category, whereas 136 iterations were required for the eight parameters selected independent of category.  No association was observed between collinearity ($\gamma$) and number of iterations or calibration.  Of the seven parameter subsets, selection independent of parameter category and equally within each category produced the largest reduction in \ac{rmse} values from the default output (starting \ac{rmse} $35.17$ mmol O$_2$ m$^{-3}$, reduction to $35.15$, $35.16$, respectively), whereas the optimal values for parameters in the Optics category had the smallest reduction ($35.17$).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 ($2.81\times 10^{-4}$, three parameters), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 ($0.03$, eight parameters, \cref{tab:calib}).

@@

% calibration table, do
%latex.default(totab[, -1], file = "", rowlabel = "Selection heuristic",     caption = cap_val, caption.loc = "top", size = "scriptsize",     label = "tab:calib", rowname = totab$heurist, rgroup = c("1",         "2", "3"), insert.bottom = foot.val, n.rgroup = c(5,         1, 1), colheads = colnms)%
\begin{table}[!tbp]
{\scriptsize
\caption{Results of model calibration to observed \ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\gamma$) and parameters of each subset are from \cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.\label{tab:calib}} 
\begin{center}
\begin{tabular}{llllll}
\hline\hline
\multicolumn{1}{l}{Selection heuristic}&\multicolumn{1}{c}{$\gamma$}&\multicolumn{1}{c}{$n$ parameters}&\multicolumn{1}{c}{mean L1}&\multicolumn{1}{c}{Iterations}&\multicolumn{1}{c}{$RMSE$*}\tabularnewline
\hline
{\bfseries 1}&&&&&\tabularnewline
~~Optics&$3.87$&$3$&$2.81\times 10^{-4}$&$7$&$35.17$\tabularnewline
~~Organic Matter&$8.61$&$5$&$3.18\times 10^{-3}$&$286$&$35.05$\tabularnewline
~~Phytoplankton&$3.97$&$7$&$0.02$&$255$&$35.03$\tabularnewline
~~Temperature&$1$&$1$&$0.02$&$3$&$35.17$\tabularnewline
~~Zooplankton&$5.2$&$4$&$0.03$&$9$&$35.17$\tabularnewline
\hline
{\bfseries 2}&&&&&\tabularnewline
~~Independent of category&$6.81$&$8$&$0.03$&$136$&$35.15$\tabularnewline
\hline
{\bfseries 3}&&&&&\tabularnewline
~~Equally within category&$5.22$&$7$&$0.03$&$105$&$35.16$\tabularnewline
\hline
\end{tabular}\end{center}}
\footnotesize *Starting \ac{rmse} $35.17$\end{table}

@@

% calibration example
\begin{figure}[!ht]

{\centering \includegraphics[width=\textwidth]{figs/calex-1}

\caption{Example of calibration steps for the phytoplankton parameter subset.  The FishTank model was calibrated using data from an approximate 12-hour flask experiment of oxygen production.  Model calibration adjusted parameters (left plots) to minimize the \ac{rmse} between experimental and modelled dissolved oxygen (top left).  The phytoplankton parameters were those identifiable for dissolved oxygen (\cref{tab:heurist1}), arranged from most (top plot) to least sensitive (bottom plot) (\cref{tab:dosens}).  The relative parameter changes in the right plots show the initial parameter values (black) and the the selected parameter values (blue) after calibration.}\label{fig:calex}
\end{figure}

% CGEM area change plots
\begin{figure}[!ht]
